Symbolic Weighted Language Models and\\ Parsing over Infinite Alphabets

### 2021-05-25 main.tex
revise SWT & SWA : no epsilon, not read tape without move for T

% fork of main.tex with more expressive models
% SWT read both tapes (under 2 heads) at every transition (even epsilon or semi-epsilon)
% SWA read tape (under head) at every transition (even epsilon)
% SW-VPA read top of stack (when not empty) at every transition

@@

% symbols
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{amsbsy}
\usepackage{bbold}
\usepackage{latexsym}
%\usepackage{amsfonts}
\usepackage{stmaryrd}
%\usepackage{mathabx}
%\usepackage{MnSymbol}
\usepackage{harmony} % simple music fonts
\usepackage{mathtools} % for arrows
%\usepackage{mathptmx}

@@

@@

%% Misc macros

\def\ie{\textit{i.e.}\xspace}
\def\eg{\textit{e.g.}\xspace}
\def\wrt{\textit{wrt}\xspace}
%\def\wlog{\textit{wlog}\xspace}
\def\etc{\textit{etc}\xspace}

@@

\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$
%the input and output alphabets~$\Sigma$ and $\Delta$ with label theory $\bar\Phi$, and the semiring $\Semiring$ 
is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Phi_{\Sigma, \Delta}$, 
respectively $\mathsf{out} : Q \to \Phi_{\Sigma, \Delta}$,
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\bar{\wei}$ is a triplet of transition functions 
$\wei_{10}$, $\wei_{01}$,  and $\wei_{11}$
%$\wei_{i, j}: Q \times Q \to \Phi_{\Sigma, \Delta}$, with $i, j \in \{ 0, 1\}$.
from $Q \times Q$ into $\Phi_{\Sigma, \Delta}$.
\end{definition}

%A pair of states $\< q, q'>$ such that $\wei_{00}(q, q') \neq \zero$
%is called an \emph{$\epsilon$-transition}.
%
For convenience, when $\wei_{ij}(q, q') = \eta \in \Phi_{\Sigma, \Delta}$, 
for $q, q' \in Q$ and $i, j \in \{ 0, 1\} \setminus \{ \<0, 0> \}$,
respectively $\mathsf{in}(q) = \eta$, $\mathsf{out}(q') = \eta$,
we shall sometimes write, overloading the function names:
$\wei_{ij}(q, a, b, q') = \eta(a, b)$ for $a \in \Sigma$,  $b \in \Delta$,
respectively $\mathsf{in}(q, a, b) = \eta(a, b)$,
$\mathsf{out}(q', a, b) = \eta(a, b)$.

\noindent 
The \SWT $T$ computes on pairs of words $\< s, t> \in \Sigma^+ \times \Delta^+$,
$s$ and $t$, being respectively called input and output word.
%
% \noindent
More precisely, the symbolic-weighted transducer~$T$ defines a mapping 
from the pairs of strings of $\Sigma^+ \times \Delta^+$ 
into~$\Semiring$,
based on the following intermediate function $\weight_T$
defined recursively for every $q, q' \in Q$, 
%$a \in \Sigma$, $b \in \Delta$ 
%for every strings of $s \in \Sigma^*$, $t \in \Delta^*$:
$a \in \Sigma$, $u \in \Sigma^*$, $b\in \Delta$, $v\in \Delta^*$,

\begin{equation}
\begin{array}[t]{lccl}
%\weight_T(q, \epsilon, \epsilon, q)  & = & \multicolumn{2}{l}{\one}\\ % or \final(q)
%\weight_T(q, \epsilon, \epsilon, q') & = & \multicolumn{2}{l}{\zero \mathrm{~if~} q \neq q'}\\
\weight_T(q, au, bv, q') & = & 
             \displaystyle\bigoplus_{q'' \in Q, u \neq \epsilon} &
    \wei_{10}(q, a, b, q'') \otimes \weight_T(q'', u, bv, q')\\
  & \oplus & \displaystyle\bigoplus_{q'' \in Q, v \neq \epsilon} &
    \wei_{01}(q, a, b, q'') \otimes \weight_T(q'', au, v, q')\\
  & \oplus & \displaystyle\bigoplus_{q'' \in Q, u,v \neq \epsilon} &
    \wei_{11}(q, a, b, q'') \otimes \weight_T(q'', u, v, q')\\
  & \oplus & \displaystyle\bigoplus_{u,v = \epsilon} &
    \wei_{11}(q, a, b, q')
\end{array}
\label{eq:SWT-weight}
\end{equation}
%
We recall that, by convention (Section~\ref{sec:semiring}), 
an empty sum with $\bigoplus$ is equal to~$\zero$. 
%
Intuitively, using a transition $\wei_{ij}(q, a, b, q')$ means for $T$:
when reading respectively $a$ and $b$ at the current positions in the input and output words, 
increment the current position in the input word if if $i = 1$, 
and in the output word if $j = 1$ (otherwise, do not change it),
and change state from $q$ to $q'$.
%
In contrast with the models of weighted transducers over finite alphabets~\cite{Mohri03ijfcs},
the input and output symbols at current positions are always read by transitions, 
even when they do not change the reading position the head's position.
This is an important feature in the case of an infinite alphabet in 
order to compare input and output symbols.
%which cannot be stored in the finite memory of the transducer.

Since $\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei_{ij}(q, a, b, q'')$ equal to $\zero$ in the above expression 
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and $b$.
This is analogous to the case of a transition's guard not satisfied by $\<a, b>$ for 
symbolic transducers.

@@

The expression of $\weight_T$ can be seen as a stateful definition of 
an edit-distance between a word $s \in \Sigma^*$ and a word $t \in \Delta^*$,
see also~\cite{Mohri03ijfcs}.
Intuitively, 
$\wei_{10}(q, a, b, r)$ is the cost of 
the deletion of the symbol~$a \in \Sigma$ in~$s$, 
$\wei{01}(q, a, b, r)$ is the cost 
of the insertion of~$b \in \Delta$ in $t$, 
and $\wei_{11}(q, a, b, r)$ is the cost 
of the substitution of  $a \in \Sigma$ by~$b \in \Delta$.
%
The cost of a sequence of such operations transforming $s$ into $t$, 
is the product, with $\otimes$, of the individual costs of the operations involved;
And the distance between $s$ and $t$ is the sum, with $\oplus$,
of all such product of costs.

\medskip\noindent
Let $\< s, t> \in \Sigma^+ \times \Delta^+$, with $s = s_1\ldots s_n$, 
and $t = t_1\ldots t_m$. 
The weight associated by $T$ to $\< s, t>$
is defined as follows:
\begin{equation}
T(s, t)  = 
\displaystyle\bigoplus_{q, q' \in Q} \mathsf{in}(q, s_1, t_1) 
\mathop{\otimes} \weight_T(q, s, t, q') \mathop{\otimes} \mathsf{out}(q', s_n, t_m)
\label{eq:SWT-value}
\end{equation}

\begin{example}
We can define with a \SWT the computation of a similarity measure 
between timed sequences
similar to dynamic time warping (DTW).

Let $\Semiring$  be the tropical (\emph{min-plus}) semiring of Figure~\ref{fig:semirings} and 
let $\Sigma = \Delta = \mathbb{R}_+$ be sets of timestamps.
We consider a \SWT with one state $q$ and transitions
$\wei_{11}(q, d, d', q) = 
 \wei_{01}(q, d, d', q) = 
 \wei_{10}(q, d, d', q) = |d' - d|$,
for all $d, d' \in \mathbb{R}_+$.
% needs reading input/output symbols even by epsilon-transitions.
The recursive definition of $\weight_T$ correspond to the dynamic programming equations of DTW
for the computation of an optimal match between words, 
the matching cost for two symbols being the 
the time distance between them.
\endex
\end{example}

@@

We consider a \SWT with two states $q_0$ and $q_1$ whose purpose 
is to compare a recorded performance $s \in \Sigma^*$
with notated music sheet $t \in \Delta^*$.
One timestamp $d_i \in \Sigma$ may corresponds 
to one notated event $\<\mathsf{e}, d'_i> \in \Sigma$, in which case 
the weight value computed by the \SWT is the time distance between both
(see transitions $\wei_{11}$ below).
%
If $\<\mathsf{e}, d'_i>$ is followed by continuations 
$\<\mathsf{-}, d'_{i+1}>$..., they are just skip with no cost (transitions $\wei_{01}$ or weight $\one$).
%transitions
\[
\begin{array}{rclcrcl}
\wei_{11}(q_0, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d| & \quad &
\wei_{11}(q_1, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d|\\
\wei_{01}(q_0, d, \< \mathsf{-}, d'>, q_0) & = & \one & &
\wei_{01}(q_1, d, \< \mathsf{-}, d'>, q_0) & = & \one\\
\wei_{10}(q_0, d, b, q_1) & = & \alpha & & \multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
Moreover, it may happen that the performers plays an extra note accidentally, but only once in a row. 
This is modelled by the transition $\wei_{10}$ with an arbitrary weight value $\alpha \in \Semiring$, 
switching from state $q_0$ (normal) to $q_1$ (error).
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances 
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and 
$\init(q_1, d, b) = \final(q_1, d, b) = \zero$, 
for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

\noindent
The \emph{Symbolic Weighted Automata} %$A = \< Q, \init, \weight, \final >$
%over $\Sigma$,  $\Semiring$ and $\bar\Phi$
are defined similarly as the transducers of Definition~\ref{def:SWT}, 
by simply omitting the output symbols.
%
In this case, the label theory $\bar\Phi$ can be reduced to a singleton $\< \Phi_\Sigma>$.
%over $\Sigma$ is reduced to
%a set $\Phi_\Sigma$ closed under~$\oplus$ and~$\otimes$.
%
\begin{definition} \label{def:SWA}
A \emph{symbolic-weighted automaton} (\SWA)
over $\Sigma$, $\Semiring$ and $\bar\Phi$
%the input alphabet~$\Sigma$ and the commutative semiring $\Semiring$ 
is a tuple
$A = \< Q, \init, {\wei_1}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Phi_\Sigma$, 
respectively $\mathsf{out} : Q \to \Phi_\Sigma,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and ${\wei_1}$ is a transition functions 
from $Q \times Q$ into $\Phi_{\Sigma}$.
\end{definition}
%      
\noindent
As above in the case of \SWT, 
when $\wei_1(q, q') = \phi \in \Phi_\Sigma$, 
respectively $\mathsf{in}(q) = \phi$, $\mathsf{out}(q') = \phi$,
we may write $\wei_1(q, a, q')$ for $\phi(a)$, 
respectively $\mathsf{in}(q, a) = \phi(a)$, $\mathsf{out}(q', a) = \phi(a)$.
%$\wei_1: Q \times \Sigma \times Q \to \Semiring$, 
The computation of $A$ on words $s \in \Sigma^+$
is defined with an intermediate function $\weight_A$, 
defined as follows for $q, q' \in Q$, $a \in \Sigma$, $u \in \Sigma^*$,
%
\begin{equation}
\begin{array}{lccl}
%\weight_T(q, \epsilon, \epsilon, q)  & = & \multicolumn{2}{l}{\one}\\ % or \final(q)
%\weight_T(q, \epsilon, \epsilon, q') & = & \multicolumn{2}{l}{\zero \mathrm{~if~} q \neq q'}\\
\weight_A(q, au, q') & = & 
            \displaystyle\bigoplus_{r \in Q, u \neq \epsilon} &
    \wei_{1}(q, a, r) \otimes \weight_A(r, u, q')\\
  & \oplus & \displaystyle\bigoplus_{u = \epsilon} & 
    \wei_{1}(q, a, q')
\end{array}
\label{eq:SWA-weight}
\end{equation}
%
\noindent
and the weight value associated by $A$ to 
$s = s_1\ldots s_n \in \Sigma^+$ is: % defined as follows:
\begin{equation}
A(s)  = 
\displaystyle\bigoplus_{q, q' \in Q} \mathsf{in}(q, s_1) 
\mathop{\otimes} \weight_A(q, s, q') \mathop{\otimes} \mathsf{out}(q', s_n)
\label{eq:SWA-value}
\end{equation}

@@

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$,
and $s \in \Sigma^+$, 
there exists an effectively constructible \SWA 
$A_{T, s}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^+$, $A_{T, s}(t) = T(s, t)$.
\end{proposition}
%
\begin{proof}
Let $T = \< Q, \init, \bar{\wei}, \final >$,
let 
$\wei: Q \times \Sigma \times \Delta \times Q \to \Semiring$
be the synthesized form of 
$\bar{\wei} = \< \wei_{10}, \wei_{01}, \wei_{11}>$,
and let $s = s_1 \ldots s_n$ with $s_1, \ldots, s_n \in \Sigma$ ($n \geq 1$).
%$n = |s|$ be the length of $s$.

The state set of $A_{T, s}$ will be $Q' = [1..n+1] \times Q$.
Its state entering weight function is defined by, 
for all $q \in Q$ and $b \in \Delta$,  
$\init'(\< 1, q>, b) = \init(q, s_1, b)$ 
and $\init'(\< i, q>, b) = \zero$ for all $1 < i \leq n+1$. 
Its state leaving weight function is defined by, for all $q \in Q$ and $b \in \Delta$,  
$\final'(\< n+1, q>, b) = \final(q, s_n, b)$, 
and $\final'(\< i, q>, b) = \zero$ for all $1 \leq i <  n+1$.

Every non-null transition of $A_{T, s}$ will
simulate a sequence of transitions of $T$ which advance in 
the input word while staying immobile in the output word, 
and then make one step in the output word (and advance in the input word or not).
%
The above initial sequence correspond to $\epsilon$-transitions of automata, 
and its total weight is computed by the following intermediate function 
$\wei'_0 : Q' \times Q' \to \Phi_{\Sigma}$, 
defined for all $q, q' \in Q$: % all $1 \leq i < n$ and $1 \leq k \leq n-i$, 
by:
\[
\begin{array}{rcl}
\wei'_0\bigl(\<i, q>, \< i, q>\bigr) & = & \one 
\quad \mathrm{if~} 1 \leq i \leq n+1,\\
%
\wei'_0\bigl(\<i, q>, \< i, q'>\bigr) & = & \zero 
\quad \mathrm{if~} 1 \leq i \leq n+1 \mathrm{~and~} q \neq q',\\
%
\wei'_0\bigl(\<i, q>, \< i+k, q'>\bigr) & = & 
\displaystyle
\bigoplus_{\begin{array}{c}
           \scriptstyle q_0,\ldots, q_k \in Q\\[-2pt]
           \scriptstyle q_0 = q, q_k = q'
           \end{array}} 
\bigotimes_{j=0}^{k-1} \bigl[\wei_{10}(q_j, q_{j+1})\bigr]_{s_{i+j}}
\quad \begin{array}[t]{l}
      \mathrm{if~} 1 \leq i < n\\
      \mathrm{and~} 1 \leq k \leq n-i.
      \end{array}
\end{array}
\]
where $\bigl[\wei_{10}(q_j, q_{j+1})\bigr]_{s_{i+j}}$ 
is the partial application $\eta_{s_{i+j}} \in \Phi_{\Delta}$
for $\eta = \wei_{10}(q_j, q_{j+1}) \in \Phi_{\Sigma, \Delta}$ ($s_{i+j} \in \Sigma$).
%
The function $\wei'_0$ is defined thanks to the closure properties of the label theory $\bar\Phi$
(Section~\ref{section:symbols}). 
% closed under partial applications 
The sum and product in its definition are finite, 
we shall see below how to compute the first sum in polynomial time.
%The effective computation of $\wei'_0$ does not require enumerating all the state sequences 
%of length $k+1$ in the sum.
%Indeed, it amounts to search short

>>>>

% SWT without epsilon-transitions
% SWT do not read the other tape in "skip" transitions 10 or 01
%     (hence this version is simpler  than V2)
% SWA without epsilon-transitions
% SW-VPA read top of stack (when not empty) at every transition 
%        (6 kinds of transitions).

@@

% symbols
\usepackage{amsmath} 
\allowdisplaybreaks
\usepackage{amssymb} 
\usepackage{amsbsy}
\usepackage{bbold}
\usepackage{latexsym}
%\usepackage{amsfonts}
\usepackage{stmaryrd}
%\usepackage{mathabx}
%\usepackage{MnSymbol}
\usepackage{harmony} % simple music fonts
\usepackage{mathtools} % for arrows
%\usepackage{mathptmx}

@@

%% allows for page break within arrays
\usepackage{longtable}

@@

%% Misc macros
\def\ie{\textit{i.e.}\xspace}
\def\eg{\textit{e.g.}\xspace}
\def\wrt{\textit{wrt}\xspace}
%\def\wlog{\textit{wlog}\xspace}
\def\etc{\textit{etc}\xspace}

@@

\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$
%the input and output alphabets~$\Sigma$ and $\Delta$ with label theory $\bar\Phi$, and the semiring $\Semiring$ 
is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$   %\Phi_{\Sigma, \Delta}
respectively $\mathsf{out} : Q \to \Semiring$  %\Phi_{\Sigma, \Delta}
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\bar{\wei}$ is a triplet of transition functions 
$\wei_{10}: Q \times Q \to \Phi_{\Sigma}$,
$\wei_{01}: Q \times Q \to \Phi_{\Delta}$, and 
$\wei_{11}: Q \times Q \to \Phi_{\Sigma, \Delta}$.
\end{definition}
%
\noindent
For convenience, we shall the following presentation of transition functions 
as functions of  
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q \to \Semiring$,
overloading the function names,
such that, for all $q, q' \in Q$, $a \in \Sigma$,  $b \in \Delta$, 
\[
\begin{array}{rcll}
\wei_{10}(q, a, \epsilon, q') & = & \phi(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi = \wei_{10}(q, q') \in \Phi_\Sigma,\\
\wei_{01}(q, \epsilon, b, q') & = & \psi(b) & 
\quad\mathrm{where~} \psi = \wei_{01}(q, q') \in \Phi_\Delta,\\
\wei_{11}(q, a, b, q') & = & \eta(a, b) & 
\quad\mathrm{where~} \eta = \wei_{11}(q, q') \in \Phi_{\Sigma, \Delta}.\\
\end{array}      
\]
%
\noindent 
The \SWT $T$ computes on pairs of words $\< s, t> \in \Sigma^* \times \Delta^*$,
$s$ and $t$, being respectively called input and output word.
%
% \noindent
More precisely, the symbolic-weighted transducer~$T$ defines a mapping 
from the pairs of strings of $\Sigma^* \times \Delta^*$ 
into~$\Semiring$,
based on the following intermediate function $\weight_T$
defined recursively for every $q, q' \in Q$, 
%$a \in \Sigma$, $b \in \Delta$ 
%for every strings of $s \in \Sigma^*$, $t \in \Delta^*$:
$a \in \Sigma$, $u \in \Sigma^*$, $b\in \Delta$, $v\in \Delta^*$, 
%
\begin{align}
\weight_T(q, \epsilon, \epsilon, q)  & = \final(q)\\
\weight_T(q, \epsilon, \epsilon, q') & = \zero \quad \mathrm{if~} q \neq q'\nonumber\\
\weight_T(q, au, \epsilon, q') & = \displaystyle\bigoplus_{q'' \in Q} 
    \wei_{10}(q, a, \epsilon, q'') \otimes \weight_T(q'', u, \epsilon, q')\nonumber\\
\weight_T(q, \epsilon, bv, q') & = \displaystyle\bigoplus_{q'' \in Q} 
    \wei_{01}(q, \epsilon, b, q'') \otimes \weight_T(q'', \epsilon, v, q')\nonumber\\
\weight_T(q, au, bv, q') & = \displaystyle\bigoplus_{q'' \in Q} 
    \wei_{10}(q, a, \epsilon, q'') \otimes \weight_T(q'', u, bv, q')\nonumber\\
                    & \oplus \displaystyle\bigoplus_{q'' \in Q} 
    \wei_{01}(q, \epsilon, b, q'') \otimes \weight_T(q'', au, v, q')\nonumber\\
                    & \oplus \displaystyle\bigoplus_{q'' \in Q} 
    \wei_{11}(q, a, b, q'') \otimes \weight_T(q'', u, v, q')\nonumber
\label{eq:SWT-weight}
\end{align}
%
We recall that, by convention (Section~\ref{sec:semiring}), 
an empty sum with $\bigoplus$ is equal to~$\zero$. 
%
Intuitively, using a transition $\wei_{ij}(q, a, b, q')$ means for $T$:
when reading respectively $a$ and $b$ at the current positions in the input and output words, 
increment the current position in the input word if and only if $i = 1$, 
and in the output word iff $j = 1$ (otherwise, do not change it),
and change state from $q$ to $q'$.
When $a = \epsilon$ (resp. $b = \epsilon$), the current symbol 
in the input (resp. output) is not read.
%
%In contrast with the models of weighted transducers over finite alphabets~\cite{Mohri03ijfcs},
%the input and output symbols at current positions are always read by transitions, 
%even when they do not change the reading position the head's position.
%This is an important feature in the case of an infinite alphabet in 
%order to compare input and output symbols.
%which cannot be stored in the finite memory of the transducer.
%
Since $\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei_{ij}(q, a, b, q'')$ equal to $\zero$ in the above expression 
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and $b$.
This is analogous to the case of a transition's guard not satisfied by $\<a, b>$ for 
symbolic transducers.

@@

The expression of $\weight_T$ can be seen as a stateful definition of 
an edit-distance between a word $s \in \Sigma^*$ and a word $t \in \Delta^*$,
see also~\cite{Mohri03ijfcs}.
Intuitively, 
$\wei_{10}(q, a, \epsilon, r)$ is the cost of 
the deletion of the symbol~$a \in \Sigma$ in~$s$, 
$\wei{01}(q, \epsilon, b, r)$ is the cost 
of the insertion of~$b \in \Delta$ in $t$, 
and $\wei_{11}(q, a, b, r)$ is the cost 
of the substitution of  $a \in \Sigma$ by~$b \in \Delta$.
%
The cost of a sequence of such operations transforming $s$ into $t$, 
is the product, with $\otimes$, of the individual costs of the operations involved;
And the distance between $s$ and $t$ is the sum, with $\oplus$,
of all such product of costs.

\medskip\noindent
%Let $\< s, t> \in \Sigma^* \times \Delta^*$, with $s = s_1\ldots s_n$, and $t = t_1\ldots t_m$. 
The weight associated by $T$ to $\< s, t> \in \Sigma^* \times \Delta^*$
is defined as follows:
\begin{equation}
T(s, t)  = 
\displaystyle\bigoplus_{q, q' \in Q} \mathsf{in}(q) 
\mathop{\otimes} \weight_T(q, s, t, q') %\mathop{\otimes} \mathsf{out}(q', s_n, t_m)
\label{eq:SWT-value}
\end{equation}

%\begin{example}
%We can define with a \SWT the computation of a similarity measure 
%between timed sequences
%similar to dynamic time warping (DTW).
%
%Let $\Semiring$  be the tropical (\emph{min-plus}) semiring of Figure~\ref{fig:semirings} and 
%let $\Sigma = \Delta = \mathbb{R}_+$ be sets of timestamps.
%We consider a \SWT with one state $q$ and transitions
%$\wei_{11}(q, d, d', q) = 
% \wei_{01}(q, d, d', q) = 
% \wei_{10}(q, d, d', q) = |d' - d|$,
%for all $d, d' \in \mathbb{R}_+$.
%% needs reading input/output symbols even by epsilon-transitions.
%The recursive definition of $\weight_T$ correspond to the dynamic programming equations of DTW
%for the computation of an optimal match between words, 
%the matching cost for two symbols being the 
%the time distance between them.
%\endex
%\end{example}

@@

We consider a \SWT with two states $q_0$ and $q_1$ whose purpose 
is to compare a recorded performance $s \in \Sigma^*$
with notated music sheet $t \in \Delta^*$.
One timestamp $d_i \in \Sigma$ may corresponds 
to one notated event $\<\mathsf{e}, d'_i> \in \Sigma$, in which case 
the weight value computed by the \SWT is the time distance between both
(see transitions $\wei_{11}$ below).
%
If $\<\mathsf{e}, d'_i>$ is followed by continuations 
$\<\mathsf{-}, d'_{i+1}>$..., they are just skip with no cost (transitions $\wei_{01}$ or weight $\one$).
%transitions
\[
\begin{array}{rclcrcl}
\wei_{11}(q_0, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d| & \quad &
\wei_{11}(q_1, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d|\\
\wei_{01}(q_0, \epsilon, \< \mathsf{-}, d'>, q_0) & = & \one & &
\wei_{01}(q_1, \epsilon, \< \mathsf{-}, d'>, q_0) & = & \one\\
\wei_{10}(q_0, d, \epsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
Moreover, it may happen that the performers plays an extra note accidentally, but only once in a row. 
This is modelled by the transition $\wei_{10}$ with an arbitrary weight value $\alpha \in \Semiring$, 
switching from state $q_0$ (normal) to $q_1$ (error).
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances 
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and 
$\init(q_1) = \final(q_1) = \zero$. 
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and 
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$, 
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

\noindent
The \emph{Symbolic Weighted Automata} %$A = \< Q, \init, \weight, \final >$
%over $\Sigma$,  $\Semiring$ and $\bar\Phi$
are defined similarly as the transducers of Definition~\ref{def:SWT}, 
by simply omitting the output symbols.
%
In this case, the label theory $\bar\Phi$ can be reduced to a singleton $\< \Phi_\Sigma>$.
%over $\Sigma$ is reduced to
%a set $\Phi_\Sigma$ closed under~$\oplus$ and~$\otimes$.
%
\begin{definition} \label{def:SWA}
A \emph{symbolic-weighted automaton} (\SWA)
over $\Sigma$, $\Semiring$ and $\bar\Phi$
%the input alphabet~$\Sigma$ and the commutative semiring $\Semiring$ 
is a tuple
$A = \< Q, \init, {\wei_1}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, %\Phi_\Sigma$, 
respectively $\mathsf{out} : Q \to \Semiring$  %\Phi_\Sigma,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and ${\wei_1}$ is a transition functions 
from $Q \times Q$ into $\Phi_{\Sigma}$.
\end{definition}
%      
\noindent
As above in the case of \SWT, 
when $\wei_1(q, q') = \phi \in \Phi_\Sigma$, 
%respectively $\mathsf{in}(q) = \phi$, $\mathsf{out}(q') = \phi$,
we may write $\wei_1(q, a, q')$ for~$\phi(a)$. 
%respectively $\mathsf{in}(q, a) = \phi(a)$, $\mathsf{out}(q', a) = \phi(a)$.
%$\wei_1: Q \times \Sigma \times Q \to \Semiring$, 
The computation of $A$ on words $s \in \Sigma^*$
is defined with an intermediate function $\weight_A$, 
defined as follows for $q, q' \in Q$, $a \in \Sigma$, $u \in \Sigma^*$,
%
\begin{align}
\weight_A(q, \epsilon, q) & = \final(q)\\
\weight_A(q, \epsilon, q') & = \zero \quad \mathrm{if~} q \neq q'\nonumber\\
\weight_A(q, au, q') & =  \displaystyle\bigoplus_{q'' \in Q} 
    \wei_{1}(q, a, q'') \otimes \weight_A(q'', u, q')\nonumber
\label{eq:SWA-weight}
\end{align}
%
\noindent
and the weight value associated by $A$ to 
$s \in \Sigma^*$ is defined as follows: %$s = s_1\ldots s_n \in \Sigma^+$ 
\begin{equation}
A(s)  = 
\displaystyle\bigoplus_{q, q' \in Q} \mathsf{in}(q) 
\mathop{\otimes} \weight_A(q, s, q') 
\label{eq:SWA-value}
\end{equation}

@@

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$,
and $s \in \Sigma^+$, 
there exists an effectively constructible \SWA 
$A_{T, s}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^+$, $A_{T, s}(t) = T(s, t)$.
\end{proposition}
%
\begin{proof}
Let $T = \< Q, \init, \bar{\wei}, \final >$,
with $\bar{\wei} = \< \wei_{10}, \wei_{01}, \wei_{11}>$,
%where the component functions of
%be the synthesized form of 
%$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q \to \Semiring$,
and let $s = s_1 \ldots s_n$ with $s_1, \ldots, s_n \in \Sigma$ ($n \geq 1$). %$n = |s|$ be the length of $s$.

The state set of $A_{T, s}$ will be $Q' = [1..n+1] \times Q$.
Its state entering weight function is defined by, 
for all $q \in Q$ $\init'(1, q) = \init(q)$ 
and $\init'(i, q) = \zero$ for all $1 < i \leq n+1$. 
Its state leaving weight function is defined by, for all $q \in Q$ 
$\final'(n+1, q) = \final(q)$, 
and $\final'(i, q) = \zero$ for all $1 \leq i <  n+1$.

Every non-null transition of $A_{T, s}$ will
simulate a sequence of transitions of $T$ performing the following steps:
advance in the input word while staying immobile in the output word, 
and then make one step in the output word (and advance in the input word or not).
%
The first steps in the sequence correspond to $\epsilon$-transitions of automata, 
and their total weight is computed by the following intermediate function 
$\wei'_0 : Q' \times Q' \to \Semiring$, 
defined for all $q, q' \in Q$: % all $1 \leq i < n$ and $1 \leq k \leq n-i$, 
by:
\[
\begin{array}{rcl}
\wei'_0\bigl(\<i, q>, \< i, q>\bigr) & = & \one 
\quad \mathrm{if~} 1 \leq i \leq n+1,\\
%
\wei'_0\bigl(\<i, q>, \< i, q'>\bigr) & = & \zero 
\quad \mathrm{if~} 1 \leq i \leq n+1 \mathrm{~and~} q \neq q',\\
%
\wei'_0\bigl(\<i, q>, \< i+k, q'>\bigr) & = & 
\displaystyle
\bigoplus_{\begin{array}{c}
           \scriptstyle q_0,\ldots, q_k \in Q\\[-2pt]
           \scriptstyle q_0 = q, q_k = q'
           \end{array}} 
\bigotimes_{j=0}^{k-1} \phi_{j, j+1}(s_{i+j})
\quad \begin{array}[t]{l}
      \mathrm{if~} 1 \leq i < n,
      \mathrm{~and~} 1 \leq k \leq n-i,\\
      \mathrm{where~} \phi_{j, j+1} = \wei_{10}(q_j, q_{j+1})
      \end{array}
\end{array}
\]
%where $\bigl[\wei_{10}(q_j, q_{j+1})\bigr]_{s_{i+j}}$ 
%is the partial application $\eta_{s_{i+j}} \in \Phi_{\Delta}$
%for $\eta = \wei_{10}(q_j, q_{j+1}) \in \Phi_{\Sigma, \Delta}$ ($s_{i+j} \in \Sigma$).
%
The function $\wei'_0$ is defined thanks to the closure properties of the label theory $\bar\Phi$
(Section~\ref{section:symbols}). 
% closed under partial applications 
The sum and product in its definition are finite, 
we shall see below how to compute the first sum in polynomial time.
%The effective computation of $\wei'_0$ does not require enumerating all the state sequences 
%of length $k+1$ in the sum.
%Indeed, it amounts to search short

\noindent
We define now the transition function 
$\wei'_1: Q' \times Q' \to \Phi_\Sigma$
of $A_{T, s}$ as follows,
for $q, q' \in Q$,
$1 \leq i \leq n$, and $0 \leq k \leq n-i$:
\[
\begin{array}{lcl}
\wei'_1\bigl(\< i, q>, \< i, q'>\bigr) & = & \wei_{01}(q, q')\\
%
\wei'_1\bigl(\< i, q>, \< i+k, q'>\bigr) & = & 
\displaystyle
\bigoplus_{\begin{array}{c}
           \scriptstyle q'' \in Q\\[-2pt]
           \scriptstyle i+k < n
           \end{array}} 
\wei'_0\bigl(\<i, q>, \< i+k, q''>\bigr) \otimes \psi %\eta_{s_{i+k}}
\quad \mathrm{where~} \psi = \wei_{01}(q'', q'),\\
 & \oplus & \displaystyle
\bigoplus_{q'' \in Q} \wei'_0\bigl(\<i, q>, \< i+k-1, q''>\bigr) \otimes \eta_{s_{i+k}}
\quad \mathrm{where~}\eta = \wei_{11}(q'', q'),\\
%
\wei'_1\bigl(\< i, q>, \< j, q'>\bigr) & = & \zero  
\quad \mathrm{if~} j < i.
\end{array}      
\]
%
In the second equation, 
$\eta_{s_{i+j}} \in \Phi_{\Delta}$ is a partial application 
for $\eta \in \Phi_{\Sigma, \Delta}$, $s_{i+j} \in \Sigma$, 
defined by $\eta_{s_{i+j}}(y) = \eta(s_{i+j}, y)$ for all $y \in \Sigma$.
